---
title: |
  | Supplementary Material for
  | Having a Ball: evaluating scoring streaks and game excitement using in-match trend estimation
author: | 
  | Claus Thorn Ekstrøm and Andreas Kryger Jensen
  | Biostatistics, Institute of Public Health, University of Copenhagen
  | ekstrom@sund.ku.dk, aeje@sund.ku.dk
date: "`r format(Sys.time(), '%d %B, %Y')`"
fontsize: 11pt  
header-includes:
  - \usepackage{bm}
  - \usepackage{amssymb}
  - \usepackage[labelfont=bf]{caption}
  - \DeclareMathOperator*{\argsup}{arg\,sup}  
  - \DeclareMathOperator*{\argmin}{arg\,min}  
  - \DeclareMathOperator*{\E}{E}
  - \DeclareMathOperator*{\Cov}{Cov}
  - \DeclareMathOperator*{\Cor}{Cor}
  - \DeclareMathOperator*{\Var}{Var}
  - \DeclareMathOperator*{\Erf}{Erf}
  - \DeclareMathOperator*{\Erfc}{Erfc}
  - \usepackage{multirow}
  - \usepackage{float}
  - \floatstyle{plaintop}
  - \restylefloat{table}
  - \usepackage[amsthm,thmmarks]{ntheorem}
  - \newtheorem{definition}{Definition}
  - \newtheorem{assumption}{Assumption}
  - \newtheorem{proposition}{Proposition}
  - \theoremstyle{nonumberplain}
  - \newtheorem{Proof}{Proof}
  - \usepackage{xcolor}
  - \newcommand{\revision}[1]{\textcolor{red}{#1}}
  - \usepackage{fontawesome5}
  - \usepackage{todonotes}
  - \newcommand{\BTheta}{\ifthenelse{\value{bbsym4theta}=1}{\text{\faBasketballBall}}{\mathbf{\Theta}}}
output:
  pdf_document: 
    keep_tex: no
    number_sections: yes
bibliography: bibliography.bib    
editor_options: 
  chunk_output_type: console
---

\newcounter{bbsym4theta}
\setcounter{bbsym4theta}{1}

# Expressions for the posterior distributions of $(d_m, d^\prime_m, d^{\prime \prime}_m)$

The joint distribution of $(d, d^\prime, d_m^{\prime \prime})$ conditional on the observed score differences $\mathcal{D}_m$ and the hyper-parameters $\BTheta_m$ evaluated at any finite vector $\mathbf{t}^\ast$ of $p$ time points is
\begin{align*}
\begin{bmatrix}d_m(\mathbf{t}^\ast)\\ d_m^{\prime}(\mathbf{t}^\ast)\\ d_m^{\prime \prime}(\mathbf{t}^\ast)\end{bmatrix} \mid \mathcal{D}_m, \BTheta_m \sim N\left(\bm{\mu},  \bm{\Sigma}\right)
\end{align*}
where $\bm{\mu} \in \mathbb{R}^{3p}$ is the column vector of posterior expectations and $\bm{\Sigma} \in \mathbb{R}^{3p \times 3p}$ is the joint posterior covariance matrix. Partitioning these as
\begin{align*}
  \bm{\mu} = \begin{bmatrix}\mu_{f}(\mathbf{t^\ast} \mid \bm{\Theta})\\ \mu_{df}(\mathbf{t^\ast} \mid \bm{\Theta})\\ \mu_{d^2\!f}(\mathbf{t^\ast} \mid \bm{\Theta})\end{bmatrix}, \quad \bm{\Sigma} = \begin{bmatrix}\Sigma_{f}(\mathbf{t^\ast},\mathbf{t^\ast} \mid \bm{\Theta}) &  \Sigma_{f,df}(\mathbf{t^\ast},\mathbf{t^\ast} \mid \bm{\Theta}) & \Sigma_{f,d^2\!f}(\mathbf{t^\ast},\mathbf{t^\ast} \mid \bm{\Theta})\\  \Sigma_{f,df}(\mathbf{t^\ast},\mathbf{t^\ast} \mid \bm{\Theta})^T & \Sigma_{df}(\mathbf{t^\ast},\mathbf{t^\ast} \mid \bm{\Theta}) & \Sigma_{df,d^2\!f}(\mathbf{t^\ast},\mathbf{t^\ast} \mid \bm{\Theta})\\ \Sigma_{f,d^2\!f}(\mathbf{t^\ast}, \mathbf{t^\ast} \mid \bm{\Theta})^T & \Sigma_{df,d^2\!f}(\mathbf{t^\ast}, \mathbf{t^\ast} \mid \bm{\Theta})^T & \Sigma_{d^2\!f}(\mathbf{t^\ast}, \mathbf{t^\ast} \mid \bm{\Theta})\end{bmatrix} 
\end{align*}
the individual components are given by
\begin{align*}
  \mu_{f}(\mathbf{t}^\ast \mid \bm{\Theta}) &= \mu_{\bm{\beta}}(\mathbf{t}^\ast) + C_{\bm{\theta}}(\mathbf{t}^\ast, \mathbf{t})\left(C_{\bm{\theta}}(\mathbf{t}, \mathbf{t}) + \sigma^2 I\right)^{-1}\left(\mathbf{Y} - \mu_{\bm{\beta}}(\mathbf{t})\right)\\
  \mu_{df}(\mathbf{t}^\ast \mid \bm{\Theta}) &= d\mu_{\bm{\beta}}(\mathbf{t}^\ast) + \partial_1 C_{\bm{\theta}}(\mathbf{t}^\ast, \mathbf{t})\left(C_{\bm{\theta}}(\mathbf{t}, \mathbf{t}) + \sigma^2 I\right)^{-1}\left(\mathbf{Y} - \mu_{\bm{\beta}}(\mathbf{t})\right) \\
  \mu_{d^2\!f}(\mathbf{t}^\ast \mid \bm{\Theta}) &= d^2\!\mu_{\bm{\beta}}(\mathbf{t}^\ast) + \partial_1^2 C_{\bm{\theta}}(\mathbf{t}^\ast, \mathbf{t})\left(C_{\bm{\theta}}(\mathbf{t}, \mathbf{t}) + \sigma^2 I\right)^{-1}\left(\mathbf{Y} - \mu_{\bm{\beta}}(\mathbf{t})\right)\\
  \Sigma_{f}(\mathbf{t}^\ast, \mathbf{t}^\ast \mid \bm{\Theta}) &= C_{\bm{\theta}}(\mathbf{t}^\ast, \mathbf{t}^\ast) - C_{\bm{\theta}}(\mathbf{t}^\ast, \mathbf{t})\left(C_{\bm{\theta}}(\mathbf{t}, \mathbf{t}) + \sigma^2 I\right)^{-1} C_{\bm{\theta}}(\mathbf{t}, \mathbf{t}^\ast)\\
  \Sigma_{df}(\mathbf{t}^\ast, \mathbf{t}^\ast \mid \bm{\Theta}) &= \partial_1\partial_2C_{\bm{\theta}}(\mathbf{t}^\ast, \mathbf{t}^\ast) - \partial_1C_{\bm{\theta}}(\mathbf{t}^\ast, \mathbf{t})\left(C_{\bm{\theta}}(\mathbf{t}, \mathbf{t}) + \sigma^2 I\right)^{-1} \partial_2C_{\bm{\theta}}(\mathbf{t}, \mathbf{t}^\ast)\\
  \Sigma_{d^2\!f}(\mathbf{t}^\ast, \mathbf{t}^\ast \mid \bm{\Theta}) &= \partial_1^2\partial_2^2 C_{\bm{\theta}}(\mathbf{t}^\ast, \mathbf{t}^\ast) - \partial_1^2 C_{\bm{\theta}}(\mathbf{t}^\ast, \mathbf{t})\left(C_{\bm{\theta}}(\mathbf{t}, \mathbf{t}) + \sigma^2 I\right)^{-1} \partial_2^2 C_{\bm{\theta}}(\mathbf{t}, \mathbf{t}^\ast)\\
  \Sigma_{f, df}(\mathbf{t}^\ast, \mathbf{t}^\ast \mid \bm{\Theta}) &= \partial_2 C_{\bm{\theta}}(\mathbf{t}^\ast, \mathbf{t}^\ast) - C_{\bm{\theta}}(\mathbf{t}^\ast, \mathbf{t})\left(C_{\bm{\theta}}(\mathbf{t}, \mathbf{t}) + \sigma^2 I\right)^{-1} \partial_2 C_{\bm{\theta}}(\mathbf{t}, \mathbf{t}^\ast)\\
  \Sigma_{f, d^2\!f}(\mathbf{t}^\ast, \mathbf{t}^\ast \mid \bm{\Theta}) &= \partial_2^2 C_{\bm{\theta}}(\mathbf{t}^\ast, \mathbf{t}^\ast) - C_{\bm{\theta}}(\mathbf{t}^\ast, \mathbf{t})\left(C_{\bm{\theta}}(\mathbf{t}, \mathbf{t}) + \sigma^2 I\right)^{-1} \partial_2^2 C_{\bm{\theta}}(\mathbf{t}, \mathbf{t}^\ast)\\
  \Sigma_{df, d^2\!f}(\mathbf{t}^\ast, \mathbf{t}^\ast \mid \bm{\Theta}) &= \partial_1 \partial_2^2 C_{\bm{\theta}}(\mathbf{t}^\ast, \mathbf{t}^\ast) - \partial_1 C_{\bm{\theta}}(\mathbf{t}^\ast, \mathbf{t})\left(C_{\bm{\theta}}(\mathbf{t}, \mathbf{t}) + \sigma^2 I\right)^{-1} \partial_2^2 C_{\bm{\theta}}(\mathbf{t}, \mathbf{t}^\ast)
\end{align*}


```{r, echo=FALSE}
rm(list=ls())
load("../analysis/est_ETIs.RData")
load("../data/nba20192020.rda")

ETIs$date <- as.Date(sapply(results, function(q) attributes(q)$date), format="%B %e, %Y")
ETIs$home <- as.factor(sapply(results, function(q) attributes(q)$home))
ETIs$away <- as.factor(sapply(results, function(q) attributes(q)$away))
```

# Median posterior excitement per team across season
```{r, echo=FALSE}
# Get team-specific average ETIs during season
team_averages <- rbind(data.frame(Team = ETIs$home, ETI = ETIs$median),
                       data.frame(Team = ETIs$away, ETI = ETIs$median))
team_summary <- t(sapply(split(team_averages, team_averages$Team), function(q) {
  c(Average = mean(q$ETI), SD = sd(q$ETI), quantile(q$ETI, 0.025), quantile(q$ETI, 0.5), quantile(q$ETI, 0.975))
}))
team_summary <- round(as.data.frame(team_summary), 2)
team_summary$Team <- rownames(team_summary)

tab <- team_summary[order(team_summary[,"Average"], decreasing=TRUE),]
knitr::kable(tab[,c("Team", "Average", "SD", "2.5%", "50%", "97.5%")], row.names=FALSE, digits=2, booktabs = TRUE, format="latex", longtable=TRUE, linesep = c("", "", "", ""), align=c("l", "r", "r|", "r", "r", "r"))
```

# Games ordered by decreasing average excitement
```{r, echo=FALSE}
library(kableExtra)
dat <- ETIs[order(ETIs$mean, decreasing=TRUE),]
res1 <- dat[, c("date", "home", "away", "mean", "lower", "median", "upper")]
colnames(res1) <- c("Date", "Home team", "Away team", "Mean ETI", "2.5% ETI", "50% ETI", "97.5% ETI")
rownames(res1) <- NULL
knitr::kable(res1, digits=2, booktabs = TRUE, format="latex", longtable=TRUE, linesep = c("", "", "", ""),  align =c("l|", "l", "l|", "r", "r", "r", "r")) %>%
  kable_styling(latex_options = c("hold_position", "repeat_header")) %>% 
  kable_styling(font_size = 8)
```

