---
title: |
  | Online Supplementary Material for
  | Having a Ball: evaluating scoring streaks and game excitement using in-match trend estimation
author: | 
  | Claus Thorn Ekstrøm and Andreas Kryger Jensen
  | Biostatistics, Institute of Public Health, University of Copenhagen
  | ekstrom@sund.ku.dk, aeje@sund.ku.dk
date: "`r format(Sys.time(), '%d %B, %Y')`"
fontsize: 11pt  
header-includes:
  - \usepackage{bm}
  - \usepackage{amssymb}
  - \usepackage[labelfont=bf]{caption}
  - \DeclareMathOperator*{\argsup}{arg\,sup}  
  - \DeclareMathOperator*{\argmin}{arg\,min}  
  - \DeclareMathOperator*{\E}{E}
  - \DeclareMathOperator*{\Cov}{Cov}
  - \DeclareMathOperator*{\Cor}{Cor}
  - \DeclareMathOperator*{\Var}{Var}
  - \DeclareMathOperator*{\Erf}{Erf}
  - \DeclareMathOperator*{\Erfc}{Erfc}
  - \usepackage{multirow}
  - \usepackage{float}
  - \floatstyle{plaintop}
  - \restylefloat{table}
  - \usepackage[amsthm,thmmarks]{ntheorem}
  - \newtheorem{definition}{Definition}
  - \newtheorem{assumption}{Assumption}
  - \newtheorem{proposition}{Proposition}
  - \theoremstyle{nonumberplain}
  - \newtheorem{Proof}{Proof}
  - \usepackage{xcolor}
  - \newcommand{\revision}[1]{\textcolor{red}{#1}}
  - \usepackage{fontawesome5}
  - \usepackage{todonotes}
  - \newcommand{\BTheta}{\ifthenelse{\value{bbsym4theta}=0}{\text{\faBasketballBall}}{\mathbf{\Theta}}}
output:
  pdf_document: 
    keep_tex: no
    number_sections: yes
bibliography: bibliography.bib    
editor_options: 
  chunk_output_type: console
---

\newcounter{bbsym4theta}
\setcounter{bbsym4theta}{1}

# Expressions for the posterior distributions of $(d_m, d^\prime_m, d^{\prime \prime}_m)$

The joint distribution of $(d, d^\prime, d_m^{\prime \prime})$ conditional on the observed score differences $\mathcal{D}_m = (\mathbf{D}_m, \mathbf{t}_m)$ and the hyper-parameters $\BTheta_m$ evaluated at any finite vector $\mathbf{t}^\ast$ of $p$ time points follows the multivariate normal distribution
\begin{align*}
\begin{bmatrix}d_m(\mathbf{t}^\ast)\\ d_m^{\prime}(\mathbf{t}^\ast)\\ d_m^{\prime \prime}(\mathbf{t}^\ast)\end{bmatrix} \mid \mathcal{D}_m, \BTheta_m \sim N\left(\bm{\mu}_m,  \bm{\Sigma}_m\right)
\end{align*}
where $\bm{\mu}_m \in \mathbb{R}^{3p}$ is the column vector of posterior means and $\bm{\Sigma}_m \in \mathbb{R}^{3p \times 3p}$ is the posterior covariance matrix. These can be partitioned as
\begin{align*}
  \bm{\mu}_m = \begin{bmatrix}\mu_{d_m}(\mathbf{t^\ast})\\ \mu_{d_m^\prime}(\mathbf{t^\ast})\\ \mu_{d_m^{\prime\prime}}(\mathbf{t^\ast})\end{bmatrix}, \quad \bm{\Sigma}_m = \begin{bmatrix}\Sigma_{d_m}(\mathbf{t^\ast},\mathbf{t^\ast}) &  \Sigma_{d_m d_m^\prime}(\mathbf{t^\ast},\mathbf{t^\ast}) & \Sigma_{d_m d_m^{\prime\prime}}(\mathbf{t^\ast},\mathbf{t^\ast})\\  \Sigma_{d_m d_m^\prime}(\mathbf{t^\ast},\mathbf{t^\ast})^T & \Sigma_{d_m^\prime}(\mathbf{t^\ast},\mathbf{t^\ast}) & \Sigma_{d_m^\prime d_m^{\prime\prime}}(\mathbf{t^\ast},\mathbf{t^\ast})\\ \Sigma_{d_m d_m^{\prime\prime}}(\mathbf{t^\ast}, \mathbf{t^\ast})^T & \Sigma_{d_m^\prime d_m^{\prime \prime}}(\mathbf{t^\ast}, \mathbf{t^\ast})^T & \Sigma_{d_m^{\prime\prime}}(\mathbf{t^\ast}, \mathbf{t^\ast})\end{bmatrix} 
\end{align*}
Using properties of the multivariate normal distribution the individual components are given by
\begin{align*}
  \mu_{d_m}(\mathbf{t}^\ast) &= \mu_{\bm{\beta}_m}(\mathbf{t}^\ast) + C_{\bm{\theta}_m}(\mathbf{t}^\ast, \mathbf{t}_m)\left(C_{\bm{\theta}_m}(\mathbf{t}_m, \mathbf{t}_m) + \sigma^2_m I\right)^{-1}\left(\mathbf{D}_m - \mu_{\bm{\beta}_m}(\mathbf{t}_m)\right)\\
  \mu_{d_m^\prime}(\mathbf{t}^\ast) &= \mu^\prime_{\bm{\beta}_m}(\mathbf{t}^\ast) + \partial_1 C_{\bm{\theta}_m}(\mathbf{t}^\ast, \mathbf{t}_m)\left(C_{\bm{\theta}_m}(\mathbf{t}_m, \mathbf{t}_m) + \sigma^2_m I\right)^{-1}\left(\mathbf{D}_m - \mu_{\bm{\beta}_m}(\mathbf{t}_m)\right) \\
  \mu_{d_m^{\prime\prime}}(\mathbf{t}^\ast) &= \mu^{\prime\prime}_{\bm{\beta}_m}(\mathbf{t}^\ast) + \partial_1^2 C_{\bm{\theta}_m}(\mathbf{t}^\ast, \mathbf{t}_m)\left(C_{\bm{\theta}_m}(\mathbf{t}_m, \mathbf{t}_m) + \sigma^2_m I\right)^{-1}\left(\mathbf{D}_m - \mu_{\bm{\beta}_m}(\mathbf{t}_m)\right)\\
  \Sigma_{d_m}(\mathbf{t}^\ast, \mathbf{t}^\ast) &= C_{\bm{\theta}_m}(\mathbf{t}^\ast, \mathbf{t}^\ast) - C_{\bm{\theta}_m}(\mathbf{t}^\ast, \mathbf{t}_m)\left(C_{\bm{\theta}_m}(\mathbf{t}_m, \mathbf{t}_m) + \sigma^2_m I\right)^{-1} C_{\bm{\theta}_m}(\mathbf{t}_m, \mathbf{t}^\ast)\\
  \Sigma_{d_m^\prime}(\mathbf{t}^\ast, \mathbf{t}^\ast) &= \partial_1\partial_2C_{\bm{\theta}_m}(\mathbf{t}^\ast, \mathbf{t}^\ast) - \partial_1C_{\bm{\theta}_m}(\mathbf{t}^\ast, \mathbf{t}_m)\left(C_{\bm{\theta}_m}(\mathbf{t}_m, \mathbf{t}_m) + \sigma^2_m I\right)^{-1} \partial_2C_{\bm{\theta}_m}(\mathbf{t}_m, \mathbf{t}^\ast)\\
  \Sigma_{d_m^{\prime\prime}}(\mathbf{t}^\ast, \mathbf{t}^\ast) &= \partial_1^2\partial_2^2 C_{\bm{\theta}_m}(\mathbf{t}^\ast, \mathbf{t}^\ast) - \partial_1^2 C_{\bm{\theta}_m}(\mathbf{t}^\ast, \mathbf{t}_m)\left(C_{\bm{\theta}_m}(\mathbf{t}_m, \mathbf{t}_m) + \sigma^2_m I\right)^{-1} \partial_2^2 C_{\bm{\theta}_m}(\mathbf{t}_m, \mathbf{t}^\ast)\\
  \Sigma_{d_m d_m^\prime}(\mathbf{t}^\ast, \mathbf{t}^\ast) &= \partial_2 C_{\bm{\theta}_m}(\mathbf{t}^\ast, \mathbf{t}^\ast) - C_{\bm{\theta}_m}(\mathbf{t}^\ast, \mathbf{t}_m)\left(C_{\bm{\theta}_m}(\mathbf{t}_m, \mathbf{t}_m) + \sigma^2_m I\right)^{-1} \partial_2 C_{\bm{\theta}_m}(\mathbf{t}_m, \mathbf{t}^\ast)\\
  \Sigma_{d_m d_m^{\prime\prime}}(\mathbf{t}^\ast, \mathbf{t}^\ast) &= \partial_2^2 C_{\bm{\theta}_m}(\mathbf{t}^\ast, \mathbf{t}^\ast) - C_{\bm{\theta}_m}(\mathbf{t}^\ast, \mathbf{t}_m)\left(C_{\bm{\theta}_m}(\mathbf{t}_m, \mathbf{t}_m) + \sigma^2_m I\right)^{-1} \partial_2^2 C_{\bm{\theta}_m}(\mathbf{t}_m, \mathbf{t}^\ast)\\
  \Sigma_{d_m^\prime d_m^{\prime\prime}}(\mathbf{t}^\ast, \mathbf{t}^\ast) &= \partial_1 \partial_2^2 C_{\bm{\theta}_m}(\mathbf{t}^\ast, \mathbf{t}^\ast) - \partial_1 C_{\bm{\theta}}(\mathbf{t}^\ast, \mathbf{t}_m)\left(C_{\bm{\theta}_m}(\mathbf{t}_m, \mathbf{t}_m) + \sigma^2_m I\right)^{-1} \partial_2^2 C_{\bm{\theta}_m}(\mathbf{t}_m, \mathbf{t}^\ast)
\end{align*}


```{r, echo=FALSE}
rm(list=ls())
load("../results/posterior_ETIs.RData")
load("../data/nba20192020.rda")

ETIs$date <- as.Date(sapply(results, function(q) attributes(q)$date), format="%B %e, %Y")
ETIs$home <- as.factor(sapply(results, function(q) attributes(q)$home))
ETIs$away <- as.factor(sapply(results, function(q) attributes(q)$away))
```

# NBA season 2019--2020 games ordered by decreasing median posterior Excitement Trend Index
```{r, echo=FALSE}
library(kableExtra)
dat <- ETIs[order(ETIs$median, decreasing=TRUE),]
res1 <- dat[, c("date", "home", "away", "mean", "lower", "median", "upper")]
colnames(res1) <- c("Date", "Home team", "Away team", "Mean", "2.5%", "Median", "97.5%")
rownames(res1) <- NULL
res1 <- cbind("No" = 1:nrow(res1), res1)
res1 <- res1[, c("No", "Date", "Home team", "Away team", "Median", "Mean", "2.5%", "97.5%")]

knitr::kable(res1, digits=2, booktabs = TRUE, format="latex", longtable=TRUE, 
             linesep = rep("", 5), 
             align = c("r", "l|", "l", "l", "r|", "r", "r", "r")) %>%
  kable_styling(latex_options = c("hold_position", "repeat_header")) %>% 
  kable_styling(font_size = 8)
```
